Abstract
\abstract{
% Parameter-efficient fine-tuning (PEFT) has emerged as a standard paradigm for adapting large models, but its direct application to real-time object detectors (\eg YOLOv3/5/6/8/9/10/11, YOLO12, RT-DETR, YOLO-World, and YOLO-Master MoE) is challenging. 
% YOLO-style detectors feature heterogeneous architectures---including dense, grouped, and depthwise convolutions, DFL layers, decoupled prediction heads, attention and open-vocabulary fusion blocks, RT-DETR decoders, and MoE experts---introducing operator, semantic, and deployment constraints that make naive PEFT target selection ineffective.

% To address this, we propose \ours, a \emph{structure-aware} PEFT framework that automates adapter placement across heterogeneous detection graphs. 
% The key insight is that effective adaptation requires respecting convolutional validity, detection-head semantics, and resource budgets, while translating generic PEFT modules into YOLO-compatible execution contracts. 
% This design enables compact and deployable adapter training without modifying standard detector interfaces for validation, checkpointing, merging, or deployment.

% %
% The framework introduces a unified planning mechanism that generalizes across architectures, ensuring stable and budget-aware adaptation. 
% %
% Extensive experiments show it achieves 
% $\bm{4.7-8.1\times}$ on-disk compression, $\bm{70\%-75\%}$ training-memory reduction, and $\bm{87.7\%}$ per-adapter distribution savings on YOLO11x, while retaining over $\bm{95\%}$ of full fine-tuning performance. 
% Ablation studies reveal that stable adapter placement is strongly architecture-conditioned rather than inherited from LLM PEFT rankings, establishing \ours{} as a practical foundation for parameter-efficient adaptation of modern real-time detectors.

% Abstract
% In few-shot, multi-domain, and resource-constrained settings, full fine-tuning of real-time object detectors such as YOLO often causes overfitting, redundant storage, and catastrophic forgetting, motivating reusable low-rank PEFT for detector adaptation. Yet PEFT methods developed for large language models do not directly transfer to real-time detection. Unlike Transformer stacks dominated by linear projections, YOLO-style detectors combine dense, grouped, and depthwise separable convolutions with Distribution Focal Loss (DFL) layers, decoupled prediction heads, and RT-DETR decoders. These heterogeneous operators, detection-specific semantics, and deployment constraints make PEFT target selection a constrained adapter placement problem.
% To address this challenge, we propose \textbf{YOLO-PEFT}, a structure-aware PEFT framework for real-time object detection. Given a detector graph, a PEFT request, and a resource budget, YOLO-PEFT builds feasible adapter configurations by filtering invalid convolutional targets, preserving detection-head semantics, assigning budget-aware ranks, and mapping generic PEFT modules onto the execution contracts required by YOLO training and deployment. This design enables compact, deployable adapter training while retaining standard interfaces for validation, checkpointing, merging, and deployment.
% Experiments show that YOLO-PEFT achieves \textbf{4.7--8.1$\times$} disk compression, reduces training memory by \textbf{70\%--75\%}, and saves \textbf{87.7\%} of per-adapter distributed training resources while retaining over \textbf{95\%} of full fine-tuning performance. Ablations further show that adapter stability is governed mainly by detector architecture, rather than stability rankings inherited from LLM-oriented PEFT methods. YOLO-PEFT thus offers a practical basis for stable, reusable, and budget-aware adaptation of modern real-time detectors.

% Abstract v260703-by linxu
Deploying YOLO-family detectors at scale demands task-, domain-, and
hardware-specific variants, yet existing adaptation is either prohibitively expensive or structurally fragile: full fine-tuning replicates and stores an entire model per variant, while PEFT techniques naively transplanted from language models can be inserted into detector layers that are structurally unsafe or semantically incompatible, causing silent accuracy collapse or convergence failure. Unlike homogeneous Transformer backbones, modern real-time detectors are heterogeneous execution graphs that interleave dense, grouped, and depthwise convolutions, multi-scale necks, decoupled detection heads, Distribution-Focal-Loss projections, attention and text-fusion modules, RT-DETR decoders, and MoE routers. Consequently,the central research question is not merely \emph{which} PEFT variant to apply, but \emph{where} an adapter can be safely and optimally placed on such a graph without compromising structural integrity or task semantics.

We propose YOLO-PEFT, a structure-aware PEFT framework that formulates detector adaptation as a constrained adapter-placement problem. Given a detector graph, a PEFT request, and a resource budget, YOLO-PEFT parses modules into operator and semantic roles, filters out structurally or semantically unsafe targets through a formal feasibility check, assigns budget-aware ranks, and lowers the resulting plan into a YOLO-compatible train--save--merge--export runtime---rejecting infeasible placements before training rather than blindly attempting them. Evaluated on PASCAL VOC object detection across YOLO11s, YOLO12s, and RT-DETR-L variants, YOLO-PEFT achieves $4.7\times$--$23.3\times$ adapter parameter compression relative to full fine-tuning, reduces training memory by $70\%$--$75\%$, and saves $87.7\%$ per-adapter distribution cost while retaining $86.1\%$--$113.2\%$ of full fine-tuning mAP50-95. Notably, PEFT can outperform full fine-tuning on small datasets by preserving backbone pretraining  knowledge ($e.g.$, YOLO12s HRA: $66.6\rightarrow74.5$ mAP50-95), yet fail catastrophically on mismatched architectures ($e.g.$, RT-DETR-L LoRA: $68.8\rightarrow59.2$ mAP50-95). Crucially, controlled ablations demonstrate that PEFT stability is \emph{architecture-conditioned} rather than method-conditioned: rankings inherited from language models fail to transfer across CNN, attention, and Transformer-decoder detectors, and unsafe placements must be rejected rather than trained. YOLO-PEFT turns YOLO adaptation from manual target-module trial and error into valid, compact, and deployable adapter planning, with an open-source runtime that supports ONNX and TensorRT export out of the box.

% \noindent\textit{Project Page:} \url{github.com/Tencent/YOLO-Master}

}
